---
title: "Classifying free days: weekends and public holidays"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Classifying free days: weekends and public holidays}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 9,
  fig.height = 5,
  dpi = 150,
  out.width = "100%"
)
library(zeitR)
library(dplyr)
```

Free-day classification sits at the heart of every chronotype metric that
zeitR computes. MSF (mid-sleep on free days), MSW (mid-sleep on workdays),
SJL (social jet lag), and CPD all depend on correctly labelling which nights
precede a free day and which do not.

By default, only **Saturdays and Sundays** are treated as free days. For
studies conducted in countries with public holidays this is usually wrong:
participants sleep differently on a Monday that follows a bank holiday
compared to a regular Monday. This vignette shows how to supply holidays, the
three supported input formats, and the mechanics of how holidays flow through
the pipeline.

---

## Holiday formats

The `holidays` parameter accepted by `run_pipeline_native()`,
`compute_sleep_metrics()`, and `compute_cpd_metrics()` understands three
forms, which can be freely combined in the same vector.

**Specific dates** — `Date` objects or `"YYYY-MM-DD"` character strings.
Use these for holidays whose calendar date changes from year to year
(Carnival, Good Friday, Easter Monday).

```{r formats-specific, eval = FALSE}
specific <- as.Date(c(
  "2019-03-04",  # Carnaval
  "2019-03-05",  # Carnaval
  "2019-04-19"   # Sexta-feira Santa
))
# character strings work too:
specific <- c("2019-03-04", "2019-03-05", "2019-04-19")
```

**Recurring holidays** — `"DD-MM"` character strings. Use these for
fixed-date holidays that fall on the same calendar day every year (Christmas,
New Year, Independence Day). The year is ignored; any date whose day and
month match is treated as a free day.

```{r formats-recurring, eval = FALSE}
recurring <- c(
  "07-09",  # Independencia do Brasil
  "12-10",  # N. Sra. Aparecida
  "02-11",  # Finados
  "15-11",  # Proclamacao da Republica
  "20-11",  # Consciencia Negra (SP municipal)
  "25-12",  # Natal
  "01-01",  # Confraternizacao Universal
  "25-01",  # Aniversario de Sao Paulo (SP municipal)
  "21-04",  # Tiradentes
  "01-05"   # Dia do Trabalho
)
```

**Mixed** — both forms in one `c()` call. This is the typical setup for a
national holiday calendar: fixed-date holidays as `"DD-MM"` and
variable-date ones as `"YYYY-MM-DD"`.

```{r formats-mixed, eval = FALSE}
sp_holidays <- c(
  # ── Recurring fixed-date holidays ────────────────────────────────────────
  "07-09", "12-10", "02-11", "15-11", "20-11",
  "25-12", "01-01", "25-01", "21-04", "01-05",
  # ── Year-specific variable-date holidays ─────────────────────────────────
  "2019-03-04", "2019-03-05",   # Carnaval 2019
  "2019-04-19"                  # Sexta-feira Santa 2019
)
```

---

## The no-holidays warning

Calling `compute_cpd_metrics()` or `compute_sleep_metrics()` without
supplying any holidays emits a warning to alert you that only weekends are
being used.

```{r warn-demo, message = TRUE, warning = TRUE}
FILE   <- system.file("extdata", "input1.txt", package = "zeitR")
TZ     <- "America/Sao_Paulo"
result <- run_pipeline_native(FILE, tz = TZ, quiet = TRUE)

# No holidays supplied — warning fires
cpd_wd_only <- compute_cpd_metrics(result$nights, tz = TZ)
```

This is intentional: a silent default that misclassifies bank holidays as
workdays is hard to catch. Once you have confirmed that weekends-only is
correct for your data (e.g. a study where participants had no common holidays),
suppress the warning globally for the session:

```{r suppress-warn}
options(zeitR.no_holidays_warn = FALSE)
```

The rest of this vignette runs with the warning suppressed.

---

## Passing holidays to the pipeline

Supply `holidays` when calling `run_pipeline_native()`. The value is stored
in `result$holidays` and does **not** affect the sleep-period detection
itself (the Vallim rules are day-of-week agnostic); it is carried through so
that downstream metric functions can pick it up automatically.

```{r pipeline}
sp_holidays <- c(
  # Recurring
  "07-09", "12-10", "02-11", "15-11", "20-11",
  "25-12", "01-01", "25-01", "21-04", "01-05",
  # Carnaval + Sexta-feira Santa 2019 (dates change each year)
  "2019-03-04", "2019-03-05", "2019-04-19"
)

result <- run_pipeline_native(
  FILE,
  tz       = TZ,
  holidays = sp_holidays,
  quiet    = TRUE
)

# holidays are stored on the result object
length(result$holidays)
head(result$holidays, 5)
```

---

## Auto-forwarding with S3 dispatch

`compute_cpd_metrics()` and `compute_sleep_metrics()` are S3 generics.
Passing a `zeitr_result` directly — rather than `result$nights` — picks up
`result$holidays` automatically as the default for the `holidays` argument.

```{r s3-dispatch}
# S3 dispatch: holidays forwarded automatically
cpd <- compute_cpd_metrics(result, tz = TZ)

# Equivalent explicit call
cpd_explicit <- compute_cpd_metrics(result$nights,
                                    tz       = TZ,
                                    holidays = result$holidays)

identical(cpd$msf_h, cpd_explicit$msf_h)
```

The same applies to `compute_sleep_metrics()`:

```{r s3-sm}
sm <- compute_sleep_metrics(result, tz = TZ)
```

If you need to override the stored holidays (e.g. to test a different
calendar), pass `holidays` explicitly and the explicit value wins:

```{r override, eval = FALSE}
cpd_alt <- compute_cpd_metrics(result,
                                tz       = TZ,
                                holidays = c("25-12", "01-01"))
```

---

## Effect on metrics

Adding public holidays to the weekend pool changes how nights are split
between the free-day and workday groups, which in turn affects MSF, MSW,
and SJL.

```{r compare}
cpd_we  <- compute_cpd_metrics(result$nights, tz = TZ)              # weekends only
cpd_all <- compute_cpd_metrics(result, tz = TZ)                     # + SP holidays

cat(sprintf("                  Weekends only   With holidays\n"))
cat(sprintf("Free days         %13d   %d\n",
            cpd_we$n_free_days, cpd_all$n_free_days))
cat(sprintf("Workdays          %13d   %d\n",
            cpd_we$n_workdays, cpd_all$n_workdays))
cat(sprintf("MSW               %13s   %s\n",
            cpd_we$msw_hms, cpd_all$msw_hms))
cat(sprintf("MSF               %13s   %s\n",
            cpd_we$msf_hms, cpd_all$msf_hms))
cat(sprintf("SJL               %10.1f min   %.1f min\n",
            cpd_we$sjl_min, cpd_all$sjl_min))
cat(sprintf("CPD               %10.1f min   %.1f min\n",
            cpd_we$cpd_min, cpd_all$cpd_min))
```

Reclassifying public holidays as free days moves nights from the workday
pool into the free-day pool. MSW and MSF shift because the composition of
each group changes; SJL reflects the true contrast between unconstrained and
constrained sleep across the whole study rather than across weekends alone.

The `compute_sleep_metrics()` output shows the same split:

```{r sm-compare}
sm_we  <- compute_sleep_metrics(result$nights, tz = TZ)
sm_all <- compute_sleep_metrics(result, tz = TZ)

data.frame(
  group        = c("Overall", "Workday", "Free day"),
  n_we         = c(sm_we$n_overall,  sm_we$n_wd,  sm_we$n_fd),
  n_all        = c(sm_all$n_overall, sm_all$n_wd, sm_all$n_fd),
  tst_wd_we    = c(NA, round(sm_we$tst_h_wd, 2), NA),
  tst_fd_we    = c(NA, NA, round(sm_we$tst_h_fd, 2)),
  tst_wd_all   = c(NA, round(sm_all$tst_h_wd, 2), NA),
  tst_fd_all   = c(NA, NA, round(sm_all$tst_h_fd, 2))
) |>
  knitr::kable(
    col.names = c("Group",
                  "N (WE only)", "N (+ holidays)",
                  "TST wd", "TST fd",
                  "TST wd", "TST fd"),
    align = "lrrrrrr",
    caption = "Night counts and mean TST (h) by group."
  )
```

---

## Batch studies spanning multiple years

For a multi-year longitudinal study, use `"DD-MM"` recurring strings for all
fixed-date national holidays and add year-specific entries for each year that
falls within the recording window.

```{r batch, eval = FALSE}
# Brazilian national holidays (recurring every year)
br_national <- c(
  "01-01",  # Confraternizacao Universal
  "21-04",  # Tiradentes
  "01-05",  # Dia do Trabalho
  "07-09",  # Independencia do Brasil
  "12-10",  # N. Sra. Aparecida
  "02-11",  # Finados
  "15-11",  # Proclamacao da Republica
  "25-12"   # Natal
)

# Variable-date holidays for each study year
variable <- c(
  # Carnaval
  "2018-02-12", "2018-02-13",
  "2019-03-04", "2019-03-05",
  "2020-02-24", "2020-02-25",
  # Sexta-feira Santa
  "2018-03-30",
  "2019-04-19",
  "2020-04-10"
)

study_holidays <- c(br_national, variable)

results <- run_pipeline_native_batch(
  "recordings/",
  tz       = TZ,
  holidays = study_holidays,
  quiet    = TRUE
)

# compute_cpd_metrics(result, tz = TZ) automatically uses each result's holidays
cpd_list <- lapply(results, compute_cpd_metrics, tz = TZ)
```

Because the `"DD-MM"` entries match any year, you only need to update
`variable` when a new study year is added — the recurring national holidays
require no changes.

---

## Summary

| Format | Example | When to use |
|---|---|---|
| `Date` / `"YYYY-MM-DD"` | `as.Date("2019-03-04")` | Holidays whose date changes year to year |
| `"DD-MM"` | `"25-12"` | Fixed-date holidays that repeat annually |
| Mixed | `c("25-12", "2019-03-04")` | Typical national calendar (most common) |

Passing `holidays` to `run_pipeline_native()` stores them on the result
object; calling `compute_cpd_metrics(result, tz = ...)` or
`compute_sleep_metrics(result, tz = ...)` with a `zeitr_result` picks them
up automatically. To silence the reminder that fires when no holidays are
supplied, use `options(zeitR.no_holidays_warn = FALSE)`.

---

## See also

- `vignette("vallim-pipeline")` — the Vallim classification rule set
- `?compute_cpd_metrics` — MSF, MSW, SJL, MSFsc, and CPD reference
- `?compute_sleep_metrics` — per-day-type sleep summary metrics
- `?run_pipeline_native` — full pipeline parameter reference
