---
title: "LIDS: ultradian sleep-cycle dynamics from actigraphy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LIDS: ultradian sleep-cycle dynamics from actigraphy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 9,
  fig.height = 4.5,
  dpi        = 150,
  out.width  = "100%"
)
library(zeitR)

has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
if (has_ggplot2) library(ggplot2)
```

Sleep isn't uniform from lights-out to wake-up: limb movement during sleep
rises and falls in ~60-110 minute cycles that track the alternation between
REM and non-REM sleep. Locomotor Inactivity During Sleep (LIDS) is a way to
recover that ultradian structure from ordinary wrist/ankle actigraphy,
without polysomnography -- turning raw movement counts into an "inactivity"
signal, then fitting a cosine to it to estimate cycle length, amplitude, and
how inactivity trends over the course of the night.

zeitR's LIDS module ports two papers:

- **Winnebeck, Fischer, Leise & Roenneberg (2018)**, *Current Biology* --
  the original method (adults/adolescents): the `100/(1+x)` transform,
  moving-average smoothing, a plain cosine fit, and the Munich Rhythmicity
  Index (MRI) for period selection.
- **Hammad, Schoch, Engelmann, Spock, Kurth & Winnebeck (2026)**, *SLEEP* --
  an infant-tuned extension: Gaussian smoothing, a *sloped* cosine (adding a
  linear trend term to capture the well-documented overnight decline in
  inactivity), and a period scan tuned for shorter (~60 min) cycles.

**This module has not yet been validated against `pyActigraphy`'s `LIDS`
class or an external reference dataset** -- see `?compute_lids` and
`?fit_lids` for exactly what's ported from which paper, and treat results
as exploratory until a parity check is run.

---

## The pipeline in one call: `compute_lids()`

For a `zeitr_result` from [`run_pipeline()`](../reference/run_pipeline.html)
or [`run_pipeline_native()`](../reference/run_pipeline_native.html),
`compute_lids()` does everything: extracts sleep bouts, transforms and fits
each one, and returns a tibble with one row per bout.

```{r pipeline-overview}
FILE <- system.file("extdata", "input1.txt", package = "zeitR")
TZ   <- "America/Sao_Paulo"

result <- run_pipeline_native(FILE, tz = TZ, quiet = TRUE)

lids_bouts <- compute_lids(result, duration_range = c(3, 12))
lids_bouts
```

Each row is one detected sleep bout (`state == 1` runs from the pipeline
output, by default -- see [Where bouts come from] below), with:

| Column | Meaning |
|---|---|
| `period_min` | Estimated ultradian cycle length (minutes) |
| `amplitude` | Oscillation amplitude (LIDS units) |
| `offset` | Inactivity level at bout start |
| `slope_per_60min` | Linear trend in inactivity across the bout |
| `phase_rad` | Phase at bout start (`0` = LIDS peak at onset) |
| `pearson_r`, `p_value`, `mri` | Fit quality and the Munich Rhythmicity Index |
| `passes_quality_filter` | `TRUE` if the bout clears the Winnebeck/Hammad quality bar |

```{r quality-summary}
mean(lids_bouts$passes_quality_filter)
summary(lids_bouts$period_min[lids_bouts$passes_quality_filter])
```

---

## Under the hood: a clean synthetic example

Because a real recording's LIDS fit is noisy, it helps to see the method on
a signal built to have a known answer first: a 60-minute sloped cosine with
a little noise, the kind of ultradian cycle Hammad et al. (2026) report in
infants.

```{r synthetic-signal}
set.seed(1)
t    <- seq(0, 300, by = 1)                                 # 5 h, 1-min epochs
true_period <- 60
lids_clean  <- 85 + 15 * cos(2 * pi * t / true_period) -
  0.05 * t + rnorm(length(t), sd = 1.5)

if (has_ggplot2) {
  ggplot(data.frame(t = t, lids = lids_clean), aes(t, lids)) +
    geom_line(colour = "#4A3F6B") +
    labs(x = "Minutes since bout onset", y = "LIDS",
         title = "Synthetic LIDS profile (period = 60 min)") +
    theme_minimal(base_size = 12)
}
```

`fit_lids()` recovers the period, amplitude, and slope directly from this
profile:

```{r fit-synthetic}
fit <- fit_lids(lids_clean, period_range = c(30, 120), period_step = 1)
fit[c("period_min", "amplitude", "offset", "slope_per_60min", "pearson_r", "mri")]
```

### From activity to LIDS: `lids_transform()`

In practice you start from raw activity counts, not a pre-built LIDS
profile. `lids_transform()` applies the `100/(1+x)` non-linear transform and
smooths the result -- higher activity means lower inactivity, and smoothing
suppresses epoch-to-epoch noise so the underlying oscillation is fittable.

```{r transform-demo}
set.seed(2)
activity <- pmax(0, 30 + 20 * sin(2 * pi * t / true_period) + rnorm(length(t), sd = 8))

lids_gaussian <- lids_transform(activity, method = "gaussian")   # Hammad et al. 2026
lids_mva      <- lids_transform(activity, method = "mva")        # Winnebeck et al. 2018

if (has_ggplot2) {
  df <- data.frame(
    t = rep(t, 2),
    lids = c(lids_gaussian, lids_mva),
    method = rep(c("gaussian", "mva"), each = length(t))
  )
  ggplot(df, aes(t, lids, colour = method)) +
    geom_line() +
    scale_colour_manual(values = c(gaussian = "#C25E2A", mva = "#4A3F6B")) +
    labs(x = "Minutes since bout onset", y = "LIDS", colour = "Smoothing",
         title = "Gaussian (Hammad 2026) vs. moving-average (Winnebeck 2018) smoothing") +
    theme_minimal(base_size = 12)
}
```

Both methods track the same underlying oscillation; Gaussian smoothing
(the default) is slightly less aggressive at suppressing sharp transients,
which matters more for the shorter cycles and higher amplitudes seen in
infants than for adults.

---

## Where bouts come from

`compute_lids()`'s `bout_source` argument controls how sleep bouts are
identified:

- **`"state"`** (used automatically above, since `result$data` already has
  a `state` column) -- contiguous `state == 1` runs from an existing zeitR
  pipeline result. This is almost always what you want if you've already
  run [`run_pipeline()`](../reference/run_pipeline.html) or
  [`run_pipeline_native()`](../reference/run_pipeline_native.html).
- **`"roenneberg"`** -- ignores any `state` column and runs the standalone
  [`detect_lids_bouts()`](../reference/detect_lids_bouts.html) relative-
  immobility detector directly on raw activity, reproducing
  Winnebeck/Hammad's own bout-detection method rather than zeitR's
  Crespo/Vallim pipelines. Use this for activity data you haven't run
  through either pipeline.
- **`"auto"`** (the default) picks `"state"` if available, `"roenneberg"`
  otherwise.

```{r bout-source-raw}
bouts <- detect_lids_bouts(
  result$data$datetime, result$data$activity,
  duration_range = c(3, 12)
)
nrow(bouts)
head(bouts)
```

---

## Quality filtering

Not every bout yields a clean ultradian fit -- a device removed mid-sleep,
an unusually short or fragmented bout, or genuine absence of rhythmicity
can all produce a poor fit. `compute_lids()` flags (rather than silently
drops) bouts failing the Winnebeck/Hammad quality bar:

```{r quality-filter}
table(lids_bouts$passes_quality_filter)

lids_bouts[!lids_bouts$passes_quality_filter,
           c("bout_id", "duration_h", "pearson_r", "p_value", "offset")]
```

Adjust the thresholds directly if the defaults (`min_r = 0.4`, `max_p =
0.05`, `offset_bounds = c(1, 99)`) don't suit your data:

```{r quality-adjust, eval = FALSE}
compute_lids(result, min_r = 0.3, max_p = 0.1)
```

---

## Study-level summaries

For a whole cohort, `study_lids_metrics()` reduces each participant's
quality-filtered bouts to one row -- median and IQR period, amplitude,
offset, and slope -- ready for `syncR::sync()` the same way
`study_sleep_metrics()` and `study_summary()` already are.

```{r study-level, eval = FALSE}
results <- run_pipeline_native_batch("recordings/", tz = "America/Sao_Paulo")
study_lids_metrics(results, min_bouts = 3)
```

---

## Next steps

- `?compute_lids` -- full argument reference, `bout_source` details, and
  the quality-filter rationale
- `?fit_lids`, `?lids_transform` -- the underlying cosine fit and
  smoothing, if you need to build a custom pipeline around them
- `?detect_lids_bouts` -- the standalone bout detector, including what
  differs from the Winnebeck/Hammad notebook it was ported from
- Winnebeck et al. (2018), *Current Biology* 28(1), 49-59.e5,
  <https://doi.org/10.1016/j.cub.2017.11.063>
- Hammad et al. (2026), *SLEEP* -- infant ultradian cycle lengthening,
  <https://zenodo.org/records/18199381>
